\chapter{Description of CD-ROM}
\label{app:cdrom}

The attached CD-ROM is written in ISO-9660 Level 1 format. Filenames
are restricted to 8+3 uppercase characters for maximum portability.

The filesystem is as follows:


\newenvironment{Ventry}[1]%
  {\begin{list}{}{\renewcommand{\makelabel}[1]{\texttt{##1}\hfil}%
    \settowidth{\labelwidth}{\textsf{#1:}}%
    \setlength{\leftmargin}{\labelwidth+\labelsep}}}%
  {\end{list}}

\begin{Ventry}{\texttt{123}}
\item[/]
	The root directory contains several `index' files. The files
	00README.TXT and 00README.HTM contain descriptions of the CD
	contents. A hypertext version of the table in appendix
	\ref{app:xcorr} can be found in XCORR.HTM (or by following the
	links in 00README.HTM).

	\begin{Ventry}{{\texttt 1234567890}}
	\item[IMG/]
		Contains images for 00README.HTM.

	\item[THESIS/]
		Postscript version of this thesis. 
		\begin{Ventry}{{\texttt 12345678}}
		\item[DOUBLE/]
			Typeset for double sided printing.
		\item[SINGLE/]
			Typeset for single sided printing.
		\end{Ventry}
		Each chapter is stored in a separate .PS file. Each
		filename starts with a two digit number denoting
		the order the files should be printed in.

	\item[WHT\_0895/]
		WHT/UES observations from August 1995 (see chapter
		\ref{ch:obs}). Observations are available as .SDF
		files and in printable postscript (.PS) format
		similar to appendix \ref{app:10lac}.

	\item[XCORR/] 
		This directory contains files used for the
		hypertext version of appendix~\ref{app:xcorr}.

		\begin{Ventry}{{\texttt 12345678}}
		\item[EPS/]
			Contains encapsulated postscript plots of the
			cross-correlation functions. The plots are
			stored in subdirectories named after the
			cross-correlation template. Files are named
			with the SWP number and this number is
			appended a C for `raw' cross-correlation
			function, and F for normalised function with
			fit. The files are compressed using
			\prog{gzip} (Lempel-Ziv) compression.

		\item[HTML/]
			Contains a hypertext summary of the results
			for each star. The files are named by the six
			digit HD number.

		\item[JPG/]
			Contains low-resolution versions of the
			postscript files above for the HTML summaries.

                \item[PSFILES/]
			Postscript versions of the hypertext summaries. Named
			by the HD catalogue number. The files are compressed
			using \prog{gzip} compression. Each file contains plots
                        of the cross-correlation functions and a summary of
			the measurements and results for each star.
		\end{Ventry}
	\end{Ventry}
\end{Ventry}



