\chapter{Line broadening parameters}
\label{app:xcorr}

The results from the cross-correlation analyses discussed in chapter
\ref{ch:xcorr} are summarised in the table on the following
pages. Column one gives the HD catalogue number, columns two and three
give spectral type and reference respectively, column four gives the
Bright Star Catalogue (BSC) spectral type. Column five shows stars
flagged as spectroscopic binaries (SB) in the BSC. A question mark
denotes a possible spectroscopic binary, 1 or 2 are for single or
double lined spectroscopic binaries respectively, and an O marks
systems where orbital data exist (all according to the BSC). Column six
gives the observation catalogue number (SWP -- Short Wavelength Prime
camera), and columns seven through ten give the corrected \vsini\
found using each of the four template spectra (see section
\ref{sec:xcorr:cal}). Velocities set in \textbf{bold} indicate the
`strongest' cross-correlation function (found from the integrated
area under the c.c.f.). Column 11 gives the adopted velocity, which,
when there are multiple spectra, is the mean of the corrected
velocites from the strongest CCFs. The spectral type references are:

\vspace{5mm}
\begin{tabular}{l@{ -- }l}
GG	& \ybcite{gg} \\
HS	& \ybcite{hs} \\
L68 	& \ybcite{l68} \\
W71	& \ybcite{wa:71supp} \\
W76	& \ybcite{wa:76} \\
\end{tabular}
\vspace{5mm}

An asterix (*) in any column indicates that a more extensive entry for
the star is to be found at the end of the table. ``\ldots'' indicates
observations that failed to give a significant cross-correlation peak,
or where the automatic routines could not complete (this would
sometimes occur if the cross-correlation peak was very narrow or
extremely wide). The hypertext table on the CD-ROM can, in some cases,
reveal why the analysis failed for a given star.

\newpage
\scriptsize
\begin{longtable}{r lcl c r rrrr r}
\multicolumn{11}{c}{Results of line broadening analysis}\\
\hline\hline
& & & \multicolumn{2}{c}{B.S.C.} & & \multicolumn{5}{c}{\vsini\ (\kms)} \\
\multicolumn{1}{c}{HD} &
\multicolumn{1}{c}{Spectral Type} & 
\multicolumn{1}{c}{Ref} & 
\multicolumn{1}{c}{Spectral Type} &
\multicolumn{1}{c}{SB} & 
\multicolumn{1}{c}{SWP} & 
$\tau$Sco & $\iota$Her & $\pi$Cet & $\alpha$Lyr & Ad.\\
\hline
\endfirsthead
\multicolumn{11}{c}{Results of line broadening analysis --- \emph{continued}}\\
\hline\hline
& & & \multicolumn{2}{c}{B.S.C.} & & \multicolumn{5}{c}{\vsini\ (\kms)} \\
\multicolumn{1}{c}{HD} &
\multicolumn{1}{c}{Spectral Type} & 
\multicolumn{1}{c}{Ref} & 
\multicolumn{1}{c}{Spectral Type} & 
\multicolumn{1}{c}{SB} & 
\multicolumn{1}{c}{SWP} & 
$\tau$Sco & $\iota$Her & $\pi$Cet & $\alpha$Lyr & Ad.\\
\hline
\endhead
\multicolumn{10}{r}{\emph{continues next page\ldots}}
\endfoot
\hline\hline
\endlastfoot
\input{tab_xcorr/appendix/tab.tex}

\end{longtable}
\normalsize

\noindent
\tref{GG -- \ybcite{gg}; HS -- \ybcite{hs}; L68 -- \ybcite{l68}; W71
-- \ybcite{wa:71supp}; -- W76 -- \ybcite{wa:76}}

\noindent
\input{tab_xcorr/appendix/comments.tex}







