\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Results of detailed spectroscopic analyses of selected early-type stars are
presented. Atmospheric parameters are derived for 23 O-type stars by
comparing high-resolution echelle observations with hydrogen/helium
non-LTE model spectra. Effective temperatures, surface gravities and
helium abundances are derived for each star, and the effects of
microturbulence on the adopted parameters are investigated.

The derived helium abundances are considered in an evolutionary
context, concentrating on trying to understand the chemically
divergent OC/ON stars. A clear relation between carbon/nitrogen
abundance anomalies and derived helium number fractions is found.
Other possible explanations for the observed abundance anomalies are
considered.

There has also been proposed (and appears to be) a relation between the
line broadening in early-type stars and their surface chemical
composition. Two possible sources for line broadening,
stellar rotational and macroturbulence, are considered and the
`best-fit' broadening type is found by careful fitting of stellar
absorption lines. Both macroturbulent and rotational velocities are
derived for the O-star sample.

A statistical analysis of the line-broadening properties of early-type
stars is also undertaken. By cross-correlating \emph{International
Ultraviolet Explorer} spectra of B stars against several stellar
template spectra, the effects of line-broadening on the
cross-correlation functions (CCFs) are investigated, and calibrations
between CCF full-width half-maximum and rotational velocity are
derived. These calibrations are used to find rotational velocities for
over 700 stars.

% Here's how to find the number of stars:
% awk '{if ($1+$2+$3+$4 != 0) {split($1,arr,"_");print arr[1];}}' 
% vsini.op.corr |sort |uniq |wc




