\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

The work done in this thesis would not have been possible without the
help of many people. I'm indebted to my supervisor, Ian Howarth, for
his expert guidance and our many discussions about astronomy. I'm
especially grateful for his careful reading of this work. His
numerous comments and suggestions have been of great help. Secondly,
Keith C. Smith has been invaluable throughout this project. The model
grids and many of the analysis codes were written by him, and I hope
that thesis shows that I at least understood a few of the things he
tried to explain to me. Thanks are also due to my second supervisor,
Raman K. Prinja, for his encouraging comments and always cheerful
attitude. A special thanks is also due to Tony Lynas-Gray for lending
me his lecture notes on stellar astrophysics and for many
drinks/dinners etc. (The list is too long to include here, but I'm
sure Tony can remember them all.) Thanks also to Serena for being my
`presence' in college\ldots without you I would never have had any
papers, Ian would never have received anything from me, and people
would have thought I'd gone missing (hang on --- they did think that
didn't they?).

This work could also not have been done (or should that be `would have
been done a lot quicker'?) without the A25/Dive Control Centre mob:
Andy (for being a very good friend), Chris E. (for asking very simple
questions which turn out to be extremely difficult to answer; MAMBO),
Matthew (for lying on the floor so I had someone to step on) and
Richard (for this \LaTeX\ style file, xpilot and ALSorts-of-stuff).

A very warm thanks to Susan for her artistic influences. I really hope
some of her talents have `leaked' into this work. I also hope that
Adrian has taught me to design and build a solid foundation for my
work (and I could always ask Andy Richardson for help if the plaster
is cracking -- or I could copy Chris `Mental/Palm-master' J. and
knock a big hole in the wall instead). Thanks to Orsola and Nic for
the use of their kitchen-shower, to the Smog-Dog for \emph{Zzzzz} `or
something', til OK, Monica og Anne for alle de oppmuntrende e-postene,
to Debs and Phil for all the cakes, chocolates and biscuits, to
Yiannis for making us wait in Warren Street station for two hours, and
to all the `little ones': Tammay, Richard P., Greg and everyone else I
can't list here.

I would also like to thank my family for all their help and support,
and especially to Nana, Grandpa, John and Kate both for their
financial support and for always being there for me.

Tusen takk til hele min norske familie, s\ae rlig Lilla og Anton, fordi
dere alltid var entusiastiske og snille, og alltid ga meg st\o tte
hvis jeg trengte det. Og selvf\o lgelig til Tia og David: Tusen takk for
alt. Dere er de beste foreldre i hele verden!

E alla fine, grazie tanto a Serena perch\'e sei la mia `Bella'.
